### 2013-03-26 doc/dissertation/document/chapters/spn_equations/spn_equations.tex
adding block jacobi preconditioning section to spn equations

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Before we can move on to solving the $SP_N$ equations with both
conventional and Monte Carlo methods, we must first understand their
spectral character to ensure the MCSA will apply. As presented in
Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
iteration matrix must be less than unity to ensure convergence and a
finite variance. Therefore, we will compute the spectral radius of the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}
Mesh element sizes were the same in all cardinal directions with
reflecting conditions on each boundary. For the energy parameter, 1
and 10 energy groups were used with upscatter/downscatter varied for
the 10 group case. For the downscatter cases, all groups could scatter
to all lower energy groups with the cross sections given in
Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
groups could scatter to all other energy groups with different cross
sections for upscatter and downscatter used. A uniform isotropic
source of neutrons is given in each group as well.

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detil of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detil of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detil of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}

\subsection{Preconditioning Strategy}
\label{subsec:spn_preconditioning}

>>>>

%%---------------------------------------------------------------------------%%
\section{Spectral Analysis of the $SP_N$ Equations}
\label{sec:spn_spectral_analysis}
Before we can move on to solving the $SP_N$ equations with both
conventional and Monte Carlo methods, we must first understand their
spectral character to ensure the MCSA will apply. As presented in
Chapter~\ref{ch:stochastic_methods}, the spectral radius of the
iteration matrix must be less than unity to ensure convergence and a
finite variance. Therefore, we will compute the spectral radius of the
$SP_N$ equations while parametrically varying the $SP_N$ order, the
$P_N$ order, the number of energy groups, and
upscattering/downscattering between energy groups. For each problem in
the parameter study, the values given in
Table~\ref{tab:spn_fixed_parameters} were fixed for each spatial
location and energy group.
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{ll}\hline\hline
      \multicolumn{1}{c}{Parameter}& 
      \multicolumn{1}{c}{Value} \\\hline
      Mesh Element Size & 1.0 \\
      Mesh Elements & $4 \times 4 \times 4$ \\
      Reflecting Boundaries & True \\
      Materials & 1 \\
      Source Strength & 1.0 \\
      Total Cross Section & 5.0 \\
      In-Group Cross Section & 0.25 \\
      Downscatter Cross Section & 1.0 \\
      Upscatter Cross Section & 0.1 \\
      Eigenvalue Solver Tolerance & $1.0\times10^{-8}$ \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Fixed parameters for the $SP_N$ spectral radius
      parameter study.} \textit{The parameters were fixed for each
      spatial location and energy group.}}
  \label{tab:spn_fixed_parameters}
\end{table}
Mesh element sizes were the same in all cardinal directions with
reflecting conditions on each boundary. For the energy parameter, 1
and 10 energy groups were used with upscatter/downscatter varied for
the 10 group case. For the downscatter cases, all groups could scatter
to all lower energy groups with the cross sections given in
Table~\ref{tab:spn_fixed_parameters} while for the upscatter case all
groups could scatter to all other energy groups with different cross
sections for upscatter and downscatter used. A uniform isotropic
source of neutrons is given in each group as well.

The matrices generated by the discretization of this problem are
sparse with a block-based form as dictated by
Eq~(\ref{eq:spn_multigroup_system}). Figure~\ref{fig:group1} gives the
sparsity pattern for a monoenergetic $SP_7$ discretization while
Figures~\ref{fig:group10ds} and \ref{fig:group10us} give the sparsity
patterns for the same problem for 10 energy groups without and with
upscatter respectively. For these figures, the parameters in
Table~\ref{tab:spn_fixed_parameters} with the number of spatial
elements reduced to $2 \times 2 \times 2$ to show the blocks generated
by the discretization.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group1.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 1-group $SP_7$
      discretization.} \textit{A $2\times 2 \times 2$ element mesh was
      used to show detail of the blocks formed by the discretization.}}
  \label{fig:group1}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group10ds.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter only.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10ds}
\end{figure}
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=3in]{chapters/spn_equations/group10us.png}
  \end{center}
  \caption{\textbf{Sparsity pattern for 10-group $SP_7$ discretization
      with downscatter and upscatter.} \textit{A $2\times 2 \times 2$ element
      mesh was used to show detail of the blocks formed by the
      discretization.}}
  \label{fig:group10us}
\end{figure}
We note a few key features of the sparsity plots. The first is that
for multigroup problems without full upscatter and downscatter
(i.e. Figure~\ref{fig:group10ds}), the resulting matrix is asymmetric
and therefore a linear solver that can handle asymmetric linear
systems is required. Nearly all problems of interest will not have
full upscattering or downscattering. Second we note the largely
diagonal character of these systems, although the blocks from
Eq~(\ref{eq:A_block_matrix}) are readily apparent. Our first attempt
at preconditioning this system will be to use the point Jacobi
preconditioning from \S\ref{sec:stochastic_preconditioning} due to
this diagonal form.

\subsection{Point Jacobi Spectral Analysis Results}
\label{subsec:spn_analysis_results}
Spectral radius computations were performed for the cases described
above for the point Jacobi preconditioned iteration matrix $\ve{H} =
\ve{I}-\ve{M}^{-1}\ve{A}$ with $\ve{M} =
diag(\ve{A})$. Table~\ref{tab:group1pj} gives the results for the
1-group case, Table~\ref{tab:group10dspj} for the 10-group case with
full downscatter only and Table~\ref{tab:group10uspj} for the 10-group
case with full downscatter and full upscatter.
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0635 & 0.6722 & 1.3144 & 1.976 \\
       & \textbf{1} & 0.0666 & 0.6728 & 1.3141 & 1.9755 \\
      $P_N$ Order & \textbf{3} & 0.0666 & 0.6822 & 1.3141 & 1.9755 \\
       & \textbf{5} & 0.0666 & 0.6822 & 1.3278 & 1.9847 \\
       & \textbf{7} & 0.0666 & 0.6822 & 1.3278 & 1.9917 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 1 energy group.}}
  \label{tab:group1pj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.0655 & 0.677 & 1.32 & 1.982 \\
       & \textbf{1} & 0.071 & 0.67777 & 1.319 & 1.982 \\
      $P_N$ Order & \textbf{3} & 0.071 & 0.687 & 1.327 & 1.9872 \\
       & \textbf{5} & 0.071 & 0.687 & 1.336 & 1.997 \\
       & \textbf{7} & 0.071 & 0.687 & 1.336 & 1.9995 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups and full
      downscatter.}}
  \label{tab:group10dspj}
\end{table}
\begin{table}[t!]
  \begin{center}
    \begin{tabular}{cccccc}\hline\hline
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{$SP_N$ Order}& 
      \multicolumn{1}{c}{}& 
      \multicolumn{1}{c}{} \\
       &   & \textbf{1} & \textbf{3} & \textbf{5} & \textbf{7}  \\
       & \textbf{0} & 0.7283 & 0.81 & 1.47 & 2.1446 \\
       & \textbf{1} & 0.7317 & 0.8 & 1.46 & 2.1368 \\
      $P_N$ Order & \textbf{3} & 0.7317 & 0.91 & 1.526 & 2.2274 \\
       & \textbf{5} & 0.7317 & 0.91 & 1.5344 & 2.2562 \\
       & \textbf{7} & 0.7317 & 0.91 & 1.5345 & 1.2842 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Spectral radius results for the point Jacobi
      preconditioned iteration matrix with 10 energy groups, full
      downscatter and full upscatter.}}
  \label{tab:group10dspj}
\end{table}
It is readily apparent from the tabulated data the point Jacobi
preconditioning is insufficient as a preconditioning mechanism for
problems of order $SP_5$ or higher. In addition, as upscatter is
introduced into the system, a much larger spectral radius is observed
at lower $SP_N$ orders, signaling a need for a better preconditioning
strategy to both ensure and improve covergence for Monte Carlo
methods.

\subsection{Preconditioning Strategy}
\label{subsec:spn_preconditioning}
If Monte Carlo methods are to be used to solve the $SP_N$ system of
equations, a different preconditioning strategy is required in order
to ensure convergence for systems of all $SP_N$ and $P_N$ orders with
arbitrary energy group structures. To achieve this, we look back to
the sparsity plots we generated in Figures~\ref{fig:group1},
\ref{fig:group10ds} and \ref{fig:group10us} as well as the multigroup
$SP_N$ equations. Initially, the diagonal character of the system led
us to try point Jacobi preconditioning with only marginal
results. From the sparsity plots we note the block structure that
ultimately arises from the multigroup scattering matrices and their
insertion into Eq~(\ref{eq:A_block_matrix}). When full upscatter and
downscatter are used the resulting blocks are completely dense while
only downscatter gives a lower triangular scattering matrix and the
block structure shown in Figure~\ref{fig:group10ds}.

Based on this both block and diagonally dominant structure for
matrices formed by the general multigroup $SP_N$ equations, we instead
choose \textit{block Jacobi} preconditioning as a left preconditioner
for the system. Like point Jacobi preconditioning, block Jacobi
preconditioning extracts the diagonal elements of the matrix as the
preconditioner where now the elements extracted are the blocks on the
diagonal as shown on the left side of
Figure~\ref{fig:block_jacobi_ex}.
\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=4in]{chapters/spn_equations/block_preconditioning}
  \end{center}
  \caption{\textbf{Block Jacobi preconditioning strategy used for the
      $SP_N$ equations.} \textit{Left: The preconditioner is formed by
      the diagonal blocks. Right: Inversion is trivial and decoupled
      by block.}}
  \label{fig:block_jacobi_ex}
\end{figure}
The inversion of this preconditioner is trivial as shown on the right
side of Figure~\ref{fig:block_jacobi_ex}. Each block can be inverted
individually and combined to form the inverse. For high performance
implementations this has several attractive properties. First, the
blocks in the matrix come from the energy/angle discretization of the
transport equation. Each block on the diagonal is bound to a mesh
element in the system (note there are 8 blocks in each of the sparsity
patterns with a mesh of $2 \times 2 \times 2$) and therefore we expect
the matrix elements forming the block to be entirely local. Second,
these blocks are typically dense and nearly lower triangular for many
transport problems meaning that established dense matrix methods based
on LU-factorization and other schemes can be used for fast inversion.
